---
title: "dde"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dde}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

``` {r echo = FALSE, results = "hide"}
output_lang <- function(x, lang) {
  writeLines(c(sprintf("```%s", lang), x, "```"))
}
output_r <- function(x) output_lang(x, "r")
output_c <- function(x) output_lang(x, "c")
knitr::opts_chunk$set(
  error = FALSE,
  fig.width = 7,
  fig.height = 5)
set.seed(1)
```

The `dde` package implements solvers for ordinary differential
equations (ODEs) and _delay_ differential equations (DDEs).  DDEs
differ from ODEs in that the right hand side depends not only on
time and the current state of the system but also on the previous
state of the system.

This seemingly innocuous dependency can create problems, especially
where the delay changes size overtime.  In particular, problems
where delays are on the order of the step size (vanishing delays)
are difficult to solve.

This package is aimed at solving non-stiff ODEs and DDEs with simple
delays.

The `deSolve` package already allows for solving delay differential
equations, though the interface and approach differs; see below for
similarities and differences.

## Ordinary differential equation models

With ODE models you will almost always be better off using
`deSolve`.  The `deTestSet` package also implements Fortran version
of the Dormand Prince algorithms here (as `deTestSet::dopri5` and
`deTestSet::dopri853`).  If you use `deSolve` then you'll have the
ability to switch between a huge number of different solvers.

The reasons to consider using `dde` over `deSolve`/`deTestSet`
would be if you

* are needing to use a DDE equation elsewhere in your package/program
* want to generate dense output for a system for later interpolation

Other than that, I would recommend using `deSolve` (which is what I
do).

For completeness, I will show how below

* R code
* Dense output and interpolation
* C code
* Exotic argument handling

### Models implemented in R

Models implemented in R look very similar to `deSolve`.  Here is
the Lorenz attractor implemented for `dde`:
``` {r }
lorenz_dde <- function(t, y, p) {
  sigma <- p$sigma
  R     <- p$R
  b     <- p$b
  y1 <- y[[1L]]
  y2 <- y[[2L]]
  y3 <- y[[3L]]
  c(sigma * (y2 - y1),
    R * y1 - y2 - y1 * y3,
    -b * y3 + y1 * y2)
}
```

The `p` argument is the parameters and can be any R object.  Here
I'll use a `list` to hold the standard Lorenz attractor parameters:
``` {r }
p <- list(sigma = 10.0,
          R     = 28.0,
          b     =  8.0 / 3.0)

tt <- seq(0, 100, length.out = 50001)
y0 <- c(1, 1, 1)
yy <- dde::dopri(y0, tt, lorenz_dde, p)
```

Here is the iconic attractor
``` {r }
par(mar=rep(.5, 4))
plot(yy[, c(2, 4)], type = "l", lwd = 0.5, col = "#00000066",
     axes = FALSE, xlab = "", ylab = "")
```

The approach above is almost identical to implementing this model
using `deSolve`:
``` {r eval = FALSE}
lorenz_ds <- function(t, y, p) {
  sigma <- 10.0
  R     <- 28.0
  b     <-  8.0 / 3.0
  y1 <- y[[1L]]
  y2 <- y[[2L]]
  y3 <- y[[3L]]
  list(c(sigma * (y2 - y1),
         R * y1 - y2 - y1 * y3,
         -b * y3 + y1 * y2))
}
yy_ds <- deSolve::ode(y0, tt, lorenz_ds, p)
```

### Dense output and interpolation

One of the nice things about the `dopri` solvers is that they do
not need to stop the integration at the times that you request
output at:
``` {r }
yy <- dde::dopri(y0, tt, lorenz_dde, p, return_statistics = TRUE)
attr(yy, "statistics")
```

Above, the number of function evaluations (~6 per step), steps, and
rejected steps is indicated (a rejected step occurs where the
solver has to reduce step size multiple times to achieve the
required accuracy).  The number of steps here is about 1/10 the
number of returned samples.  This works because the solver here
returns "**dense output**" which allows it to interpolate the
solution between points that it has not visited.  This is supported
by many of the solvers in `deSolve`, too.

In contrast with `deSolve`, the dense output here can be collected
and worked with later, though doing this requires a bit of faff.

Specify the history length; this needs to be an _overestimate_
because once the end of the history buffer is reached it will be
silently overwritten to return the *last* steps in history.  (This
is the behaviour required to support delay models without running
out of memory).
``` {r }
yy2 <- dde::dopri(y0, range(tt), lorenz_dde, p, return_minimal = TRUE,
                  n_history = 5000, return_history = TRUE)
```

With these arguments `yy2` is a 3 x 1 matrix, but it comes with a
massive "history" matrix":
``` {r }
dim(yy2)
h <- attr(yy2, "history")
dim(h)
```

The contents of this matrix are designed to be opaque (i.e., I may
change how things are represented at a future time).  However, the
solution can be interpolated to any number of points using this
matrix:
``` {r }
yy2 <- dde::dopri_interpolate(h, tt)
all.equal(yy2, yy[, -1], check.attributes = FALSE)
```

## Delay differential equation models

Implementing a delay differential equation model (vs an ODE model)
means that you refer to the model state at a previous point in
time.  To do that, you use the the `ylag` function, of which `dde`
provides interfaces in both R and C.

### A model in R

This is a simple SEIR (Susceptible - Exposed - Infected -
Resistant) model from epidemiology.  Once exposed to the disease,
an individual exists in an "Exposed" state for exactly 14 days
before becoming "Infected" (you could model this with a series of
compartments and get a distribution of exposed times).
``` {r }
seir <- function(t, y, p) {
  b <- 0.1
  N <- 1e7
  beta <- 10.0
  sigma <- 1.0 / 3.0
  delta <- 1.0 / 21.0
  t_latent <- 14.0
  Births <- N * b
  surv <- exp(-b * t_latent)

  S <- y[[1L]]
  E <- y[[2L]]
  I <- y[[3L]]
  R <- y[[4L]]

  tau <- t - t_latent
  y_lag <- dde::ylag(tau, c(1L, 3L)) # Here is ylag!
  S_lag <- y_lag[[1L]]
  I_lag <- y_lag[[2L]]

  new_inf <- beta * S * I / N
  lag_inf <- beta * S_lag * I_lag * surv / N

  c(Births  - b * S - new_inf + delta * R,
    new_inf - lag_inf - b * E,
    lag_inf - (b + sigma) * I,
    sigma * I - b * R - delta * R)
}
```

The model needs to know how many susceptible individuals there were
14 days ago, and how many infected there were 14 days ago.  To get
this from the model, we use

```r
y_lag <- dde::ylag(tau, c(1L, 3L))
```

to get the values of the first and third variables (S and I) at time
`tau`.  Alternatively you can get all values with

```r
y_lag <- dde::ylag(tau)
```

or get them individually

```r
S_lag <- dde::ylag(tau, 1L)
I_lag <- dde::ylag(tau, 3L)
```

The `ylag` function can only be called from within an integration;
it will throw an error if you try to call it otherwise.

What happens when we start though?  If time starts at 0, then the
first `tau` is -14 and we have no history then.  `dde` keeps track
of the initial state of the system and if a time before this is
requested it returns the initial state of a variable.  This is
going to be reasonable for many applications but will lead to
discontinuities in the *derivative* of your solution (and the
second derivative and so on).  This can make the problem hard to
solve, and it may be preferable to provide your own information
(see the deSolve implementation below for one possible way of
implementing this).

To integrate the problem, use the `dde::dopri` function (which by
default will use the 5th order method, which is probably the best
bet for most problems).  You need to provide arguments:

* `n_history`: number of history elements to retain.  If this is
  too low then the integration will stop with an error and you can
  increase it
* `return_history`: set this to `FALSE` if you won't want the
  history matrix returned; returning it costs a little time and if
  you don't want to inspect it it's better to leave it off
``` {r }
y0 <-  y0 <- c(1e7 - 1, 0, 1, 0)
tt <- seq(0, 365, length.out = 100)
yy <- dde::dopri(y0, tt, seir, NULL, n_history = 1000L, return_history = FALSE)
matplot(tt, yy[, 2:5], type="l")
```

## Differences with deSolve

deSolve has a function `dede` that implements a delay differential
equation solver, supporting solutions using `lsoda` and other
solvers.  `dde` differs in both approach and interface and these
are documented here for users familiar with `deSolve`.  This
section is not needed for basic use of the package, but may be
useful if you have used deSolve, especially with compiled or DDE
models.

By default the delayed variables are computed using interpolation
of the solution using Hermitian (cubic) interpolation along the
time dimension.  This works surprisingly well, but we found that
`lsoda` and other solvers got confused on some large problems
(~2000 equations, 3 delays), possibly because the order of accuracy
of the interpolated solution is much lower than the accuracy of the
actual problem.  This manifested in the solver locking up in a
matrix algebra routine involved with approximating the Jacobian of
the solution.  The package `PBSddesolve`, based on `solv95`, takes
a similar approach and may have similar limitations.

The `dde` solver uses the "dense output" that the Dormand-Prince
solvers generate; this means that the value of lagged variables can
be immediately looked up without any additional interpolation, and
that the accuracy of the lagged variables will be the same as the
integrated variables.

* more flexible handling of the parameters object (which is not global)
* output only hit using interpolation
* R functions do not return lists, and return output via an
  attribute (this may change?)

### Models implemented in R

Above, I implemented a derivative function for an SEIR model for `dde` as
``` {r echo=FALSE, results="asis"}
output_r(capture.output(print(seir)))
```

The implementation using `deSolve` looks very similar:
``` {r }
seir_deSolve <- function(t, y, parms) {
  b <- 0.1
  N <- 1e7
  beta <- 10
  sigma <- 1 / 3
  delta <- 1 / 21
  t_latent <- 14.0
  I0 <- 1
  Births <- N * b
  surv <- exp(-b * t_latent)

  S <- y[[1L]]
  E <- y[[2L]]
  I <- y[[3L]]
  R <- y[[4L]]

  tau <- t - t_latent
  if (tau < 0.0) { # NOTE: assuming that t0 is always zero
    S_lag <- parms$S0
    I_lag <- parms$I0
  } else {
    y_lag <- deSolve::lagvalue(tau, c(1L, 3L))
    S_lag <- y_lag[[1L]]
    I_lag <- y_lag[[2L]]
  }

  new_inf <- beta * S * I / N
  lag_inf <- beta * S_lag * I_lag * surv / N

  list(c(Births  - b * S - new_inf + delta * R,
         new_inf - lag_inf - b * E,
         lag_inf - (b + sigma) * I,
         sigma * I - b * R - delta * R))
}
```

The differences are that:

* `deSolve` requires that the derivatives are returned as a list,
  whereas `dde` uses a numeric vector (see below for details about this)
* `deSolve` requires that you provide the initial values for the
  lagged values (and we also need to know what the initial *time*
  is too, but I'm assuming that as zero)
* The appropriate function for pulling previous values from the
  history buffer is `deSolve::lagvalue` (for `dde` it is
  `dde::ylag`)

Aside from this the code is essentially identical.

To run the model with `deSolve`, use `deSolve::dede` which
automatically sets up a history buffer of 10000 elements (the
`mxhist` element of the control list alters this).
``` {r }
y0 <-  y0 <- c(1e7 - 1, 0, 1, 0)
tt <- seq(0, 365, length.out = 100)
initial <- list(S0 = y0[[1]], I0 = y0[[3]])
yy_ds <- deSolve::dede(y0, tt, seir_deSolve, initial)
```

This produces output that the same as `dde`:
``` {r }
yy_dde <- dde::dopri(y0, tt, seir, NULL, n_history = 1000L,
                     return_history = FALSE)
op <- par(mfrow=c(1, 2), mar=c(4, .5, 1.4, .5), oma=c(0, 2, 0, 0))
matplot(tt, yy_dde[, -1], type="l", main = "dde")
matplot(tt, yy_ds[, -1], type="l", main = "deSolve", yaxt="n")
```

The performance of both packages is fairly similar, taking a few
tens of milliseconds to run on my machines
``` {r }
tR <- microbenchmark::microbenchmark(times = 30,
  deSolve = deSolve::dede(y0, tt, seir_deSolve, initial),
  dde = dde::dopri(y0, tt, seir, NULL, n_history = 1000L,
                   return_history = FALSE))
tR
```

### Models implemented in C

``` {r include = FALSE}
.dlls <- local({
  build <- sprintf("%s.c", c("seir", "seir_ds"))
  files <- file.path(dde:::dde_example_path(), build)
  lapply(files, dde:::shlib, "dde_")
})
```

The compiled code interface for `deSolve` has greatly influenced
`dde` and models implemented in either framework will be similar.
Eventually `dde` may support a fully `deSolve` compatible interface
but for now there are a few differences.

``` {r echo = FALSE, results = "asis"}
output_c(readLines(system.file("examples/seir_ds.c", package = "dde")))
```

This looks very similar to the `dde` version above but:

* `parms` (or whatever parameters are called) are handled as a
  global variable that is updated via a model initialisation
  function, whereas in `dde` they're passed in as a `void` pointer
* We need to keep track of the initial state of the system via
  passing in `t0` and initial conditions for `S` and `I` * There is
  an argument `double *yout` for additional output variables (of
  length `*ip`; in `dde` these are handled via a separate function.
* The lagvalue function must be explicitly defined (which requires
  loading some R-related headers (in `dde` this is achieved by
  including `<dde/dde.h>` and `<dde/dde.c>`.

Apart from these details, the model definition should appear very
similar.
``` {r }
initial <- c(0.0, y0[[1]], y0[[3]])

zz_ds <- deSolve::dede(y0, tt, "seir_deSolve", initial,
                       initfunc = "seir_initmod", dllname = "dde_seir_ds")
zz_dde <- dde::dopri(y0, tt, "seir", numeric(), dllname = "dde_seir",
                     n_history = 1000L, return_history = FALSE)
```

Check that outputs of these models are the same as the R version
above:
``` {r }
all.equal(zz_ds, yy_ds, check.attributes = FALSE)
all.equal(zz_dde, yy_dde, check.attributes = FALSE)
```

Here, the timings are even closer and have dropped from on the
order of 20 milliseconds to 0.5 milliseconds; so we're getting a
~40x speed up from using compiled code.
``` {r }
tC <- microbenchmark::microbenchmark(
  deSolve = deSolve::dede(y0, tt, "seir_deSolve", initial,
                          initfunc = "seir_initmod", dllname = "dde_seir_ds"),
  dde = dde::dopri(y0, tt, "seir", numeric(), dllname = "dde_seir",
                   n_history = 1000L, return_history = FALSE))
tC
```

The difference in speed will tend to increase as the models become
larger (in terms of numbers of equations and parameters).  On the
other hand, constructing large models in C can be a hassle (but see
[odin](https://mrc-ide.github.io/odin) for a possible solution).

You can extract a little more performance by tweaking options to
`dde::dopri`; in particular, adding `return_minimal = TRUE`
will avoid transposing the output, binding the times on, and (if
given) avoiding binding output variables.  These costs may be
nontrivial with bigger models, though the cost of running a larger
model will likely be larger still.  Previous version of R suffered
from a large cost of looking up the address of the compiled
function (Windows may still take longer to do this than macOS/Linux).
In that case, use `getNativeSymbolInfo("seir")` and pass that
through to `dopri` as the `func` argument.
``` {r }
ptr <- getNativeSymbolInfo("seir")
tC2 <- microbenchmark::microbenchmark(
  deSolve = deSolve::dede(y0, tt, "seir_deSolve", initial,
                          initfunc = "seir_initmod", dllname = "dde_seir_ds"),
  dde = dde::dopri(y0, tt, "seir", numeric(), dllname = "dde_seir",
                   n_history = 1000L, return_history = FALSE),
  dde2 = dde::dopri(y0, tt, ptr, numeric(), n_history = 1000L,
                    return_history = FALSE, return_minimal = TRUE))
tC2
```


``` {r include = FALSE}
for (x in .dlls) {
  tryCatch({
    dyn.unload(x$dll)
    unlink(x$tmp, recursive = TRUE)
  }, error = function(e) NULL)
}
```
